White Dwarf Mergers on Adaptive Meshes\\ II. Collisions

### 2016-01-29 papers/wdmerger_II/paper.tex
Update discussion of the burning limiters to reflect new AMR approach

In \castro, the reactions are coupled to the hydrodynamics using Strang splitting.
In a given timestep advance $\Delta t$, we first evolve the reactions alone through
a time interval $\Delta t / 2$. Then, we evolve the hydrodynamics for $\Delta t$,
and we evolve the reactions again for a further $\Delta t / 2$. The principal
drawback of this approach is that the reactions and the hydrodynamics can become
decoupled from each other. A common solution to this problem presented in
the literature has been to limit the size of the timestep and thereby limit the
extent of this decoupling \citep{raskin:2010,hawley:2012}, which we adopt here.
(A related approach presented in \citet{kushnir:2013} is to limit the magnitude
of the energy release in any given integration by scaling all the reaction rates.)
At the start of each advance, we limit the size of the timestep to be the smaller
of the minimum hydro timestep (limited by the CFL condition), and the minimum burning
timestep. In a given zone, the burning timestep is defined as:
\begin{equation}
  \Delta t_{\text{burning}} = f_b\, \frac{e}{\dot{e}}.
\end{equation}
Here $\dot{e}$ is an estimate of the time rate of change of the internal energy
from nuclear burning. We estimate this using the energy change in the last burning
update, divided by the length of the burning timestep. The factor $f_b$ then
determines by what fraction we would like to allow the internal energy to change
in the current timestep (assuming that $\dot{e}$ does not change quickly from
timestep to timestep, which is not always the case). By making $f_b$ smaller,
we can control the magnitude of the decoupling between the reactions and the hydro.
A typical choice in the literature is $f_b \sim 0.3$. This timestep limiting has
now been implemented in \castro\ and the factor $f_b$ can be set at runtime.

>>>>

In \castro, the reactions are coupled to the hydrodynamics using Strang splitting.
In a given timestep advance $\Delta t$, we first evolve the reactions alone through
a time interval $\Delta t / 2$. Then, we evolve the hydrodynamics for $\Delta t$,
and we evolve the reactions again for a further $\Delta t / 2$. The principal
drawback of this approach is that the reactions and the hydrodynamics can become
decoupled from each other. A common solution to this problem presented in
the literature has been to limit the size of the timestep and thereby limit the
extent of this decoupling \citep{raskin:2010,hawley:2012}, which we adopt here 
and have implemented in \castro. Defining the nuclear energy injection timescale 
$t_e$:
\begin{equation}
  t_e \equiv \frac{e}{|\dot{e}|},
\end{equation}
where $\dot{e}$ is an estimate of the time rate of change of the internal energy
from nuclear burning, we define a burning-limited timestep $\Delta t_{b}$:
\begin{equation}
  \Delta t_{b} = f_{\Delta t}\, t_e.
\end{equation}
We estimate $\dot{e}$ using the energy change in the last burning
update, divided by the length of the burning timestep. The factor $f_{\Delta t}$ then
determines by what fraction we would like to allow the internal energy to change
in the current timestep (assuming that $\dot{e}$ does not change quickly from
timestep to timestep, which is not always the case). By making $f_{\Delta t}$ smaller,
we can control the magnitude of the decoupling between the reactions and the hydro.
We choose by default $f_{\Delta t} = 0.1$, a choice which is on par with others in 
the literature. The sensitivity of results to this choice will be discussed in 
\autoref{sec:2D:timestep}. The factor $f_{\Delta t}$ can be set at runtime in \castro.

At the start of each advance, we limit the size of the timestep to be the smaller
of the minimum hydro timestep (limited by the CFL condition), and the minimum of the
burning timestep across all zones.

\citet{kushnir:2013} point out this is not the only way for the burning to
become numerically unstable; another failure mode is when the energy injection timescale
$t_e$ is shorter than the sound-crossing time $t_s$ in a zone.
The goal is to ensure that the following condition holds:
\begin{equation}
  t_s \leq f_{\Delta x} t_e \label{eq:burning_limiter_2}
\end{equation}
The sound crossing time, $t_s$, is given by $\Delta x / c_s$, 
where $c_s$ is the sound speed and $\Delta x$ is the (minimum) 
zone width. The parameter $f_{\Delta x}$ then determines the minimum
ratio of the nuclear energy generation timescale to the 
sound-crossing time. \citet{kushnir:2013} choose $f_{\Delta x} = 0.1$ 
for their simulations, and we do too (this parameter can be set 
at runtime in \castro).

\citet{kushnir:2013} implemented this criterion by artificially 
limiting the magnitude of the energy release after a burn. As 
this is not physical, we take a different approach. Since we 
cannot directly control the energy injection timescale, we 
must find a way to alter the sound-crossing timescale. 
We can achieve this by adding levels of refinement in 
regions that do not satisfy \autoref{eq:burning_limiter_2},
which effectively lowers the $\Delta x$ and thus the
sound-crossing time. We keep tagging zones for refinement
based on this criterion until the criterion is satisfied
on the finest level. Since the concern is regions that 
may detonate, we also tag nearby zones in a buffer region
which do not themselves satisfy the criterion,
so that a detonation in a single timestep cannot 
escape into non-refined regions. The width of the buffer 
region should thus be at least as large as the number of 
timesteps before a regridding procedure is performed.
We choose a value of two for both the number of zones in the 
buffer region and the number of steps in between regrids,
for all simulations in this paper.

While this approach may add several AMR levels 
to a simulation, we agree with \citet{kushnir:2013} that 
solving this numerical instability is crucial to avoiding
early, unphysical detonations, and so it is unavoidable
if a correct evaluation of the burning phase is desired.
A simulation that does not solve this problem in this way
(or some analogous manner) will not obtain the correct amount 
of burning, and will not converge properly with resolution. 
Fortunately, the regions where this criterion is not satisfied 
are fairly localized on the grid, so the amount of additional 
computational work is not exceedingly large.
